### 2019-10-11 ch1_introduction/content.tex
update

The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}. Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).

@@

During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.

@@

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The naturally occurring cell-to-cell variability happens at the level of gene expression. Gene expression itself can be controlled at different levels (Figx), one of which is gene regulation by transcription.
%Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.

Fig. x: \textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e. splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity.
%From: (https://www.ncbi.nlm.nih.gov/books/NBK26885/figure/A1220/?report=objectonly) Meer inhoudelijke info: https://www.ncbi.nlm.nih.gov/books/NBK26885/ Ref: Alberts B, Johnson A, Lewis J, Raff M, Roberts K, Walter P. An Overview of Gene Control. InMolecular Biology of the Cell. 4th edition 2002. Garland Science.

@@

Transcription factors (TFs) modulate the rate of gene transcription by binding and recruiting the transcriptional machinery to \emph{cis}-regulatory regions (enhancers, and silencers) that are typically located in the promotor region of target genes. These bindings may result in increased or decreased gene expression. 
There are several TF families of which members share structural characteristics (e.g. zink finger, helix-loop-helix).

Many TFs are commonly present in virtually all cell types (e.g. NF-κB), while others are specific for cells and developmental stages. Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic. Moreover, the production of a specific molecule might require several gene regulatory cascades. Studying the active parts of a cell's GRN can thus reveal which dynamic processes are taking place within a cell.
% Ref: Lambert SA, Jolma A, Campitelli LF, Das PK, Yin Y, Albu M, Chen X, Taipale J, Hughes TR, Weirauch MT. The human transcription factors. Cell. 2018 Feb 8;172(4):650-65.
% The kappa in NF-κB wordt niet goed omgezet in pdf doc

@@

IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation. The use of different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously. Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation). While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.

Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells. Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies.

@@

\subsubsection{Omics}
On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\mycomment{Demonstrate the masking effect of bulk analyses.}

\subsubsection{Single-cell omics}
Transformative technological advances in microvolume sequencing allowed Tang et al. to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A). During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells (Figure \ref{fig:profiling_convergence}B) and measuring other levels of information such as proteomic expression levels (Figure \ref{fig:profiling_convergence}C).

\begin{figure}[htb!]
  \centering
  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
  % also figure with cell numbers
  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
  \caption{
    A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}. 
    B. Cell numbers vs. gene count over time.
    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
    \fbox{\color{red} TODO: combine images, rewrite the description.}
  }
  \label{fig:profiling_convergence}
\end{figure}

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the Human Cell Atlas. The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells.

\section{Computational tools}
The new types of analyses permitted by single-cell omics harbour exceptional opportunities to discover new aspects of biology and redefine existing knowledge (Figure \ref{fig:comp_tools}A). These include the following categories\cite{zappia_exploringsinglecellrnaseq_2018}.
\begin{itemize}
	\item \textbf{Data imputation}: \ldots.
	\item \textbf{Data integration}: \ldots.
	\item \textbf{Clustering}: \ldots.
	\item \textbf{Dimensionality reduction}: providing a visual and informative overview of a given dataset.
	\item \textbf{Trajectory inference}: identifying and characterising transitions between different cellular states.
	\item \textbf{Trajectory alignment}: \ldots.
	\item \textbf{Network inference}: inferring gene regulatory interactions between transcription factors across individual cells.
\end{itemize}
\mycomment{To do: write a section for each of these categories.}

Developing new computational tools to perform these analyses has proven challenging for two main reasons\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}. Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity. In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}.
    \fbox{\color{red} TODO: combine images, rewrite the description.}
  }
  \label{fig:comp_tools}
\end{figure}

@@

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
  \caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
  \label{fig:bulk_vs_singlecell}
\end{figure}

@@

@@

We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends. This includes help speed up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes). Several of these package receive millions of downloads per year (devtools, remotes, rlang).

>>>>

\subsection{Central dogma}
The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}. Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).

@@

% KDP: je zegt niets over epigenetica, die min of meer bepaald welke gebieden in het DNA worden vertaald (dit leidt je misschien te ver...)
During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.

@@

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The naturally occurring cell-to-cell variability happens at the level of gene expression. Gene expression itself can be controlled at different levels (Figure~\ref{fig:controlexpression}), one of which is gene regulation by transcription.

\begin{figure}[htb!]
	\centering\includegraphics[width=.8\linewidth]{fig/control_gene_expression} % TODO: change image
	\caption{
		\textbf{Levels of controlling gene expression} can happen at the level of transcription, RNA processing (i.e. splicing), RNA transport and localization, mRNA translation, mRNA degradation and protein activity \cite{alberts_overviewgenecontrol_2002}.
	}
	\label{fig:controlexpression}
\end{figure}

% KDP: niet alleen TFs maar volledige epigenetica regelt of een gen al dan niet tot expressie wordt gebracht, dus misschien kort vermelden dat het toch iets complexer is

@@

Transcription factors (TFs) modulate the rate of gene transcription by binding and recruiting the transcriptional machinery to \emph{cis}-regulatory regions (enhancers, and silencers) that are typically located in the promotor region of target genes. These bindings may result in increased or decreased gene expression. 
There are several TF families of which members share structural characteristics (e.g. zinc finger, helix-loop-helix).

Many TFs are commonly present in virtually all cell types (e.g. NF-$\kappa$B), while others are specific for cells and developmental stages\cite{lambert_humantranscriptionfactors_2018}. Typically, the same TF can regulate the rate of transcription of many target genes in different cell types, indicating that these gene regulatory networks (GRNs) are dynamic. Moreover, the production of a specific molecule might require several gene regulatory cascades. Studying the active parts of a cell's GRN can thus reveal which dynamic processes are taking place within a cell.

% KDP: misschien interessant om over tumor heterogeniteit te spreken: waarvan we vroeger dachten dat homogene massa van cellen was, blijkt toch veel complexer te zijn --> nu mogelijk om met sc technieken te bestuderen?

@@

% KDP: antigen-antibody: eerder over protein spreken ipv over antigen dan wordt het toch iets duidelijker, denk ik: een antibody herkent en bindt een bepaald proteine/eiwit,...
IHC (and many other biotechnologies) visualises antigen-antibody reactions by attaching particular molecules to the antibody, such as an enzyme that catalyses a colour-producing reaction, or a fluorescent chemical compound that can re-emit light upon light excitation. The use of different colours (wavelengths) allows measuring expression levels of different antibodies simultaneously. Characterising cells in a semi-quantifiable way is labour intensive, however; since it involves acquiring an image of many cells and drawing a contour around each cell (called cell segmentation).
While modern implementations of IHC improve the throughput drastically by using robots to automate the image acquisition and computer software to automate cell segmentation, the procedure is still labour intensive as the robots and computer software still needs to be kept in check.

Flow cytometry\cite{fulwyler_electronicseparationbiological_1965} is a technique which circumvents imaging and segmentation issues by having a steady stream of cells run through a laser and measuring the amount of light scattered from those cells.
% KDP: circumvents segmentation: if you want to investigate tissues you have to dissociate the tissue 
Flow cytometry technology enables to measure protein expression levels for millions of cells and tens of different antibodies. 
% KDP: vermeld hier hoeveel merkers flow cytometrie tegenwoordig aankan, want je komt er later op terug

@@

\subsubsection{Omics}
On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting\cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner.

There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively.

Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.

Traditionally, profiling nano-litre volumes was infeasible. To capture enough material to generate a profile, an ensemble of cells needs to be pooled and lysed together, thereby granting the technology the name "bulk" omics. 
Bulk omics is a major workhorse in <...> and has applications in <...>. To capture enough material, an ensemble of cells is pooled and profiled all together.

\mycomment{TODO: fill in}

A notable downside of bulk omics arises when the cells are biomolecularly heterogeneous (Figure~\ref{fig:bulk_vs_sc}A). Since the profile is a population average (or rather, a summation), information pertaining the individual profiles of cell subpopulations cannot be discerned (Figure~\ref{fig:bulk_vs_sc}B).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\LARgefigure]{fig/bulk_vs_sc} 
	\caption{
		\textbf{The 'masking' effect of bulk omics.} 
		\textbf{A:} Cells from several subpopulations are incorrectly assumed to be homogeneous and are profiled with a single bulk omics experiment.
		\textbf{B:} The signals from the different subpopulations are masked. The resulting profile is dissimilar from the majority of cells it is supposed to represent.
	}
	\label{fig:bulk_vs_sc}
\end{figure}

\subsubsection{Single-cell omics}
Transformative technological advances in microvolume sequencing allowed Tang et al. to analyse the transcriptome at single-cell resolution\cite{tang_mrnaseqwholetranscriptomeanalysis_2009}, thereby bringing single-cell biology and omics together to create single-cell omics (Figure \ref{fig:profiling_convergence}A). During the decade that followed, the number of single-cell omics technologies has skyrocketed, allowing to profile tens of thousands of cells and measuring other levels of information (e.g. proteomic expression levels and spatial location).
% KDP: wat bedoel je met: such as proteomic expr levels

\begin{figure}[htb!]
  \centering
  \includegraphics[width=\linewidth]{fig/singlecell_omics_evolution} 
  \caption{
    \textbf{} %Convergence of "Omics" Biology and Single-Cell Biology. Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}. 
  }
  \label{fig:profiling_convergence}
\end{figure}

The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge. Some of these opportunities lie in efforts like the Human Cell Atlas. The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories connecting the different cell types. As part of this endeavour, the consortium will likely profile the whole transcriptomes tens or even hundreds of millions of cells. 
% KDP: hier heb je het al eens over gehad: zou je dit niet samenbrengen?

\section{Computational tools}
Whole-genome profiling at single-cell level allows new types of analyses with which to study cellular heterogeneity at a hitherto unseen throughput. 
The new types of analyses permitted by single-cell omics present several computational challenges\cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}. This necessitates the development of novel computational tools, either because the problem statement of the performed analysis is completely novel, or to adapt existing methodology to new data characteristics (e.g. dimensionality and noise levels). Over the past five years, already 450 new tools for analysing single-cell omics data have been developed\cite{zappia_exploringsinglecellrnaseq_2018}. These tools typically fall within one of nine categories (Figure \ref{fig:comp_tools}). We discuss each category in the following subsections.

\mycomment{Todo: describe the characteristics of single cell omics data in the subsection on 'single cell omics'. -> counts data. additional information from the cell can be present, such as the time of sampling in a time-series experiment or a cell type if the cells have been sorted using a FACS experiment. information on the features are typically restricted to the gene symbol, or an identifier of a xxx, such as ensembl or entrez. in comparison to bulk data, single-cell omics data is typically very sparse, due to technical and biological stochasticity. if several transcripts of a low abundance gene are not captured, this will be seen as a dropout. dropouts can also be biological, due to transcriptional bursting. }

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/comptools/comptools_3.pdf} 
  \caption{
  	\textbf{Computational tools developed for processing single-cell omics data usually lie within one of nine categories.}
  	\textbf{A:} Normalisation, separating meaningful biological signal from technical noise or irrelevant biological information.
  	\textbf{B:} Data imputation, masking dropouts.
  	\textbf{C:} Data integration, combining multiple single-cell modalities.
  	\textbf{D:} Dimensionality reduction, providing a visual overview of the main cellular heterogeneity in the dataset.
  	\textbf{E:} Clustering, grouping together similar cells.
  	\textbf{F:} Trajectory inference, identifying and characterising transitions between different cellular states.
  	\textbf{G:} Network inference, predicting gene regulatory interactions from cell-to-cell heterogeneity.
  	\textbf{H:} RNA velocity, inferring the directionality of trajectories by predicting the future state of each cell.
  	\textbf{I:} Trajectory alignment, aligning multiple trajectories.
  }
  \label{fig:comp_tools}
\end{figure}

\subsection{Normalisation}

\subsection{Data imputation}

\subsection{Data integration}

@@

\subsection{Clustering}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
  \caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: Already used this figure in the previous section
  \label{fig:bulk_vs_singlecell}
\end{figure}

@@

% mention differential networks

\subsection{RNA velocity}

\subsection{Trajectory alignment}

@@

We also helped maintain or extend other packages on Github, CRAN or Bioconductor on which our software depends. This includes speeding up parts of the dependency (slingshot), adding new functionality (devtools, ParamHelpers), fixing bugs (proxyC, rlang, monocle, splatter, slingshot), becoming a maintainer of orphaned packages (diffusionMap, princurve, GillespieSSA), and extending the documentation (devtools, mlr, remotes). Several of these package receive millions of downloads per year (devtools, remotes, rlang).
